---
title: "Displaying nestedLogit models as LaTeX equations"
author: "Michael Friendly"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  message   = FALSE,
  warning   = FALSE,
  comment   = "#>"
)
```

# Setup

```{r load}
devtools::load_all()   # load dev version (not yet installed)
library(equatiomatic)
data(Womenlf, package = "carData")

comparisons <- logits(work = dichotomy("not.work",
                                       working = c("parttime", "fulltime")),
                      full = dichotomy("parttime", "fulltime"))

wlf.nested <- nestedLogit(partic ~ hincome + children,
                           dichotomies = comparisons,
                           data = Womenlf)
```

# Baseline: equations for individual sub-models

`extract_eq()` already works on the individual `glm` sub-models. These serve as
the baseline for comparison.

```{r submodel-work}
mod.work <- models(wlf.nested, "work")
extract_eq(mod.work)
```

Note that the response is rendered as `..y` — the internal variable name used
when fitting the sub-model — rather than as a meaningful label.

```{r submodel-full}
mod.full <- models(wlf.nested, "full")
extract_eq(mod.full)
```

# Equations for the full `nestedLogit` object

`extract_eq(wlf.nested)` dispatches to `extract_eq.nestedLogit`, which applies
`extract_eq()` to each sub-model and replaces `..y` with the dichotomy name.

The result is a named list of class `"nestedLogit_equations"`.

```{r nestedLogit-eqns}
eqns <- extract_eq(wlf.nested)
class(eqns)
names(eqns)
```

## Work dichotomy (not working vs. working)

```{r eqn-work}
eqns$work
```

## Full-time dichotomy (part-time vs. full-time)

```{r eqn-full}
eqns$full
```

# Inspect raw LaTeX

```{r raw-latex}
cat("work:\n", as.character(eqns$work), "\n\n")
cat("full:\n", as.character(eqns$full), "\n")
```

# With actual coefficient values

`extract_eq()` accepts a `use_coefs = TRUE` argument to substitute the
fitted coefficient values into the equation.

```{r eqn-coefs}
eqns_coef <- extract_eq(wlf.nested, use_coefs = TRUE)
eqns_coef$work
eqns_coef$full
```

# Gators data: 2 dichotomies, continuous predictor

The `gators` data models alligator food choice (Other, Fish, Invertebrates) as
a function of body length.  The two dichotomies use `logits()` rather than
`continuationLogits()`, and the only predictor is continuous.

```{r gators-model}
data(gators)
gators.dichots <- logits(
  other    = dichotomy("Other", c("Fish", "Invertebrates")),
  fish_inv = dichotomy("Fish", "Invertebrates")
)

# view the dichotomies
gators.dichots

gators.nested <- nestedLogit(food ~ length,
                             dichotomies = gators.dichots,
                             data = gators)
```

Display the model equations:
```{r gators-eqns}
eqns_gators <- extract_eq(gators.nested)
eqns_gators$other
eqns_gators$fish_inv
```

Use the estimated coefficients:
```{r gators-eqns-coef}
eqns_gators_coef <- extract_eq(gators.nested, use_coefs = TRUE)
eqns_gators_coef$other
eqns_gators_coef$fish_inv
```

# GSS data: 3 dichotomies via `continuationLogits()`, interaction term

The `GSS` data models educational attainment (`degree`) across four ordered
levels using continuation logits.  The predictor includes an interaction
(`parentdeg * year`), giving a more complex RHS and three dichotomy equations.

```{r gss-model}
data(GSS)
gss.nested <- nestedLogit(degree ~ parentdeg * year,
                           dichotomies = continuationLogits(
                             c("<highschool", "highschool", "college", "graduate")
                           ),
                           data = GSS)
names(models(gss.nested))
```

```{r gss-eqns}
eqns_gss <- extract_eq(gss.nested)
eqns_gss[[1]]
eqns_gss[[2]]
eqns_gss[[3]]
```

# Testing `submodel` argument and `knit_print`

## `submodel`: extract a single equation by name

`submodel = "work"` should return a bare `"equation"` object that renders
directly — equivalent to `extract_eq(wlf.nested)$work` but without the
`$` indexing.

```{r submodel-arg}
extract_eq(wlf.nested, submodel = "work")
```

```{r submodel-arg-coef}
extract_eq(wlf.nested, submodel = "full", use_coefs = TRUE)
```

Error on a bad name:

```{r submodel-bad, error=TRUE}
extract_eq(wlf.nested, submodel = "bad_name")
```

## `submodel` with other arguments

`submodel` and `...` arguments can be combined:

```{r submodel-coef}
extract_eq(wlf.nested, submodel = "work", use_coefs = TRUE)
```

```{r submodel-colors}
extract_eq(wlf.nested, submodel = "work",
           greek_colors = c("black", "blue", "blue"),
           var_colors   = c(hincome = "red", children = "darkgreen"))
```

## Gators: `submodel` with underscore name

```{r submodel-gators}
extract_eq(gators.nested, submodel = "other")
extract_eq(gators.nested, submodel = "fish_inv")
```
