---
title: "Displaying nestedLogit models as LaTeX equations"
author: "Michael Friendly"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  message   = FALSE,
  warning   = FALSE,
  comment   = "#>"
)
```

# Setup

```{r load}
devtools::load_all()   # load dev version (not yet installed)
library(equatiomatic)
data(Womenlf, package = "carData")

comparisons <- logits(work = dichotomy("not.work",
                                       working = c("parttime", "fulltime")),
                      full = dichotomy("parttime", "fulltime"))

wlf.nested <- nestedLogit(partic ~ hincome + children,
                           dichotomies = comparisons,
                           data = Womenlf)
```

# Baseline: equations for individual sub-models

`extract_eq()` already works on the individual `glm` sub-models. These serve as
the baseline for comparison.

```{r submodel-work}
mod.work <- models(wlf.nested, "work")
extract_eq(mod.work)
```

Note that the response is rendered as `..y` — the internal variable name used
when fitting the sub-model — rather than as a meaningful label.

```{r submodel-full}
mod.full <- models(wlf.nested, "full")
extract_eq(mod.full)
```

# Equations for the full `nestedLogit` object

`extract_eq(wlf.nested)` dispatches to `extract_eq.nestedLogit`, which applies
`extract_eq()` to each sub-model and replaces `..y` with the dichotomy name.

The result is a named list of class `"nestedLogit_equations"`.

```{r nestedLogit-eqns}
eqns <- extract_eq(wlf.nested)
class(eqns)
names(eqns)
```

## Work dichotomy (not working vs. working)

```{r eqn-work}
eqns$work
```

## Full-time dichotomy (part-time vs. full-time)

```{r eqn-full}
eqns$full
```

# Inspect raw LaTeX

```{r raw-latex}
cat("work:\n", as.character(eqns$work), "\n\n")
cat("full:\n", as.character(eqns$full), "\n")
```

# With actual coefficient values

`extract_eq()` accepts a `use_coefs = TRUE` argument to substitute the
fitted coefficient values into the equation.

```{r eqn-coefs}
eqns_coef <- extract_eq(wlf.nested, use_coefs = TRUE)
eqns_coef$work
eqns_coef$full
```
