---
title: "Other Examples of Nested Logit Models"
author: "Michael Friendly and John Fox"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.height = 5,
  fig.width  = 7,
  comment   = "#>"
)
```

# Introduction

This document explores datasets beyond those already packaged with `nestedLogit`
(`Womenlf`, `GSS`, `gators`, `HealthInsurance`) that could profit from treatment
as nested logit models, either because:

- the response categories have a natural **hierarchical tree structure** and the
  independence-of-irrelevant-alternatives (IIA) assumption of plain multinomial
  logit is implausible across nests, or
- the response is **ordinal** and **continuation logits** (a special case of
  nested dichotomies) are a natural model.

The best candidates identified are:

| Dataset | Package | Response levels | Structure | N |
|---------|---------|-----------------|-----------|---|
| `Chile` | carData | A/N/U/Y | two competing trees | 2700 |
| `TravelMode` | AER | air/train/bus/car | IIA violation | 210 |
| `Fishing` | mlogit | beach/pier/boat/charter | shore vs. vessel | 1182 |
| `Arthritis` | vcd | None/Some/Marked | ordinal continuation | 84 |
| `survey$Smoke` | MASS | Never/Occas/Regul/Heavy | ordinal continuation | 237 |
| `pneumo` | VGAM | normal/mild/severe | ordinal continuation | 8 |

```{r load-packages}
library(nestedLogit)
library(nnet)       # multinom()
library(car)        # Anova()
library(ggplot2)
```

---

# Chilean plebiscite voting intent (`carData::Chile`) {#chile}

## Data

The 1988 Chilean national plebiscite asked citizens whether Pinochet should
remain in power for another eight years. Voting intention was recorded as:

- **Y** (yes, support Pinochet)
- **N** (no, oppose Pinochet)
- **U** (undecided)
- **A** (abstain)

`Chile` is in `carData`, which is already a dependency of `nestedLogit`, so
no extra package is needed.

```{r chile-data}
data("Chile", package = "carData")

# Drop the 168 rows with missing vote so all models use the same cases
chile <- Chile[!is.na(Chile$vote), ]
nrow(chile)
table(chile$vote)
```

The strong predictor `statusquo` is a scale measuring attitude toward the
political status quo (higher = more supportive of Pinochet's regime).

```{r chile-eda}
# statusquo distribution by vote
boxplot(statusquo ~ vote, data = chile,
        xlab = "Vote", ylab = "Attitude toward status quo",
        main  = "Chile 1988: status quo attitude by voting intention")
```

## Two competing tree structures

The key theoretical question is the **ordering of the decision process**.

### Engagement-first tree

The voter first decides whether to participate meaningfully (cast a Y or N
vote) vs. disengage (abstain or remain undecided). Engagement is then governed
by demographic/civic variables; direction by political attitude.

```
          vote
         /    \
     engaged  disengaged
      /   \     /    \
     Y     N   A      U
```

```{r chile-dichots-eng}
dichots_eng <- logits(
  engage    = dichotomy(engaged    = c("Y", "N"),
                        disengaged = c("A", "U")),
  direction = dichotomy("Y", "N"),
  disengage = dichotomy("A", "U")
)
dichots_eng
```

### Direction-first tree

The voter first forms a political opinion (pro or anti status quo) and only
then decides whether to act on it. `statusquo` should dominate the direction
split; demographics should govern the two engagement splits.

```
         vote
        /    \
      pro    anti
     /   \   /   \
    Y     U N     A
```

```{r chile-dichots-dir}
dichots_dir <- logits(
  direction  = dichotomy(pro  = c("Y", "U"),
                         anti = c("N", "A")),
  engage_pro = dichotomy("Y", "U"),
  engage_ant = dichotomy("N", "A")
)
dichots_dir
```

## Fitting the models

```{r chile-fit}
f <- vote ~ statusquo + age + sex + education + income

# Multinomial logit baseline (reference = "N")
chile2 <- within(chile, vote <- relevel(vote, ref = "N"))
chile_multi <- multinom(f, data = chile2, trace = FALSE)

# Nested logit: engagement-first tree
chile_eng <- nestedLogit(f, dichotomies = dichots_eng, data = chile)

# Nested logit: direction-first tree
chile_dir <- nestedLogit(f, dichotomies = dichots_dir, data = chile)
```

```{r chile-summaries}
summary(chile_eng)
summary(chile_dir)
```

### Omnibus tests

```{r chile-anova}
Anova(chile_eng)
Anova(chile_dir)
```

## Which tree fits better?

Both nested models have identical parameter counts (5 predictors × 3 sub-models
= 15 slopes plus 3 intercepts), so AIC is a fair comparison.

```{r chile-aic}
data.frame(
  model   = c("Multinomial", "Engagement tree", "Direction tree"),
  logLik  = c(logLik(chile_multi), logLik(chile_eng), logLik(chile_dir)),
  AIC     = c(AIC(chile_multi),    AIC(chile_eng),    AIC(chile_dir))
)
```

The direction-first tree should fit better if `statusquo` is primarily a
predictor of political opinion (Y/N vs. U/A) rather than of engagement. We can
check this by inspecting the `statusquo` coefficients in each sub-model of the
engagement tree: they should be large in `direction` and small in `engage` and
`disengage`.

## Predicted probabilities

```{r chile-newdata}
new_chile <- data.frame(
  statusquo = seq(-1.5, 1.5, by = 0.25),
  age       = median(chile$age,    na.rm = TRUE),
  sex       = factor("F",  levels = levels(chile$sex)),
  education = factor("S",  levels = levels(chile$education)),
  income    = median(chile$income, na.rm = TRUE)
)
```

```{r chile-predict}
pred_eng <- as.data.frame(predict(chile_eng, newdata = new_chile))
pred_dir <- as.data.frame(predict(chile_dir, newdata = new_chile))
```

```{r chile-plot, fig.height=5, fig.width=9}
# as.data.frame() returns long format: one row per (observation × response level)
# with predictor columns, plus 'response', 'p', 'se.p', 'logit', 'se.logit'
pred_long <- rbind(
  transform(pred_eng, model = "Engagement-first tree"),
  transform(pred_dir, model = "Direction-first tree")
)

ggplot(pred_long, aes(x = statusquo, y = p, colour = response)) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~ model) +
  labs(x      = "Attitude toward status quo",
       y      = "Predicted probability",
       colour = "Vote",
       title  = "Chile 1988: predicted vote probabilities by tree structure") +
  theme_bw()
```

```{r chile-plot-logit, fig.height=5, fig.width=6}
# Logit-scale plot using the new scale= argument
plot(chile_eng, "statusquo",
     others = list(age       = median(chile$age, na.rm = TRUE),
                   sex       = "F",
                   education = "S",
                   income    = median(chile$income, na.rm = TRUE)),
     scale  = "logit",
     xlab   = "Attitude toward status quo",
     main   = "Engagement-first tree")
```

---

# Travel mode choice (`AER::TravelMode`) {#travel}

```{r travel-pkg, include=FALSE}
have_AER <- requireNamespace("AER",   quietly = TRUE)
have_dplyr <- requireNamespace("dplyr", quietly = TRUE)
```

This section requires the `AER` and `dplyr` packages.

```{r travel-eval, eval=have_AER && have_dplyr}
library(AER)
library(dplyr)
data("TravelMode", package = "AER")
```

## Data structure

`TravelMode` is in **long format**: 840 rows = 210 individuals × 4 travel
alternatives (air, train, bus, car). Variables such as `wait` and `gcost` are
*alternative-specific* — they differ across modes for the same individual.

`nestedLogit()` requires **one row per individual** with the chosen mode as the
response. Only **individual-specific** predictors (`income`, `size`) work here
directly; alternative-specific costs would require a different approach.

```{r travel-reshape, eval=have_AER && have_dplyr}
tm <- TravelMode |>
  filter(choice == "yes") |>
  select(mode, income, size) |>
  mutate(mode = relevel(factor(mode), ref = "car"))

table(tm$mode)
```

## Nesting structure

The canonical nested structure separates *private* (car) from *public* transit,
then within public separates *air* from *ground*, then *train* from *bus*.
This reflects the IIA problem: adding a near-identical alternative (e.g., a
second bus service) should not equally reduce the probability of choosing car.

```{r travel-dichots, eval=have_AER && have_dplyr}
travel_dichots <- logits(
  pvt_pub  = dichotomy("car", public = c("air", "train", "bus")),
  air_grnd = dichotomy("air", ground = c("train", "bus")),
  tr_bus   = dichotomy("train", "bus")
)
travel_dichots
```

## Fitting

```{r travel-fit, eval=have_AER && have_dplyr}
tm_multi  <- multinom(mode ~ income + size, data = tm, trace = FALSE)
tm_nested <- nestedLogit(mode ~ income + size,
                         dichotomies = travel_dichots,
                         data = tm)
summary(tm_nested)
Anova(tm_nested)
```

## Predicted probabilities

```{r travel-predict, eval=have_AER && have_dplyr}
new_tm <- data.frame(income = seq(10, 60, by = 10), size = 1)

# as.data.frame() returns long format; reshape to wide for comparison
pred_tm_nested_long <- as.data.frame(predict(tm_nested, newdata = new_tm))
nested_wide <- reshape(
  pred_tm_nested_long[, c("income", "size", "response", "p")],
  idvar     = c("income", "size"),
  timevar   = "response",
  direction = "wide"
)
names(nested_wide) <- sub("^p\\.", "", names(nested_wide))

pred_tm_multi <- as.data.frame(predict(tm_multi, newdata = new_tm,
                                       type = "probs"))

cat("Nested logit:\n")
cbind(income = new_tm$income, round(nested_wide[, levels(tm$mode)], 3))

cat("Multinomial logit:\n")
cbind(income = new_tm$income, round(pred_tm_multi[, levels(tm$mode)], 3))
```

## Model comparison

```{r travel-aic, eval=have_AER && have_dplyr}
data.frame(
  model  = c("Multinomial", "Nested"),
  logLik = c(logLik(tm_multi), logLik(tm_nested)),
  AIC    = c(AIC(tm_multi),    AIC(tm_nested))
)
```

---

# Fishing mode choice (`mlogit::Fishing`) {#fishing}

```{r fishing-pkg, include=FALSE}
have_mlogit <- requireNamespace("mlogit", quietly = TRUE)
```

This section requires the `mlogit` package.

```{r fishing-load, eval=have_mlogit}
data("Fishing", package = "mlogit")
```

## Overview

1182 US recreational anglers each chose one of four fishing modes:
`beach`, `pier`, `boat`, `charter`. The natural nesting separates
*shore-based* from *vessel-based* alternatives:

```
       fishing
      /        \
   shore      vessel
  /     \    /      \
beach  pier boat  charter
```

Shore-based modes share unobserved attributes (no boat required, lower cost
and commitment) that vessel-based modes do not, so IIA is plausible within
but not across groups.

`Fishing` is in long format (4 rows per individual) with alternative-specific
`price` and `catch` variables. As with `TravelMode`, only the individual-
specific variable `income` can be used directly with `nestedLogit()`.

```{r fishing-reshape, eval=have_mlogit}
fishing_wide <- Fishing[Fishing$mode == Fishing$mode[1], ]  # placeholder
# TODO: proper reshape to one-row-per-person with chosen mode as response
# The mlogit package stores data differently; reshaping requires care.
# head(Fishing)
```

The `mlogit`-specific long format requires reshaping before `nestedLogit()`
can be applied. This is left as a TODO pending a clean data preparation step.

---

# Arthritis treatment outcome (`vcd::Arthritis`) {#arthritis}

```{r arthritis-pkg, include=FALSE}
have_vcd <- requireNamespace("vcd", quietly = TRUE)
```

```{r arthritis-load, eval=have_vcd}
data("Arthritis", package = "vcd")
str(Arthritis)
table(Arthritis$Improved)
```

## Overview

84 patients in a double-blind clinical trial received either an active treatment
or a placebo. The outcome `Improved` is *ordinal*:

```
      Improved
     /        \
  None      Any improvement
            /             \
         Some           Marked
```

Continuation logits are natural here: dichotomy 1 asks whether the patient
improved at all; dichotomy 2 asks, given improvement, whether it was marked
rather than merely some. Treated patients should have higher probabilities at
both transitions.

```{r arthritis-dichots, eval=have_vcd}
arth_dichots <- continuationLogits(c("None", "Some", "Marked"))
arth_dichots
```

```{r arthritis-fit, eval=have_vcd}
arth_nested <- nestedLogit(Improved ~ Treatment + Sex + Age,
                           dichotomies = arth_dichots,
                           data = Arthritis)
summary(arth_nested)
Anova(arth_nested)
```

```{r arthritis-plot, eval=have_vcd}
plot(arth_nested, x.var = "Age",
     others = list(Treatment = "Treated", Sex = "Female"),
     xlab   = "Age",
     main   = "Arthritis: Treatment = Treated, Sex = Female")
```

---

# Smoking status (`MASS::survey`) {#smoking}

```{r survey-load}
data("survey", package = "MASS")
# Drop NAs on Smoke
sv <- survey[!is.na(survey$Smoke), ]
table(sv$Smoke)
```

## Overview

237 Australian university students. `Smoke` is ordinal:
`Never` < `Occas` < `Regul` < `Heavy`.

Continuation logits:
- Dichotomy 1: Never vs. ever smoked
- Dichotomy 2: Occasional vs. regular+
- Dichotomy 3: Regular vs. heavy

```{r smoke-dichots}
smoke_dichots <- continuationLogits(c("Never", "Occas", "Regul", "Heavy"))
smoke_dichots
```

```{r smoke-fit}
smoke_nested <- nestedLogit(Smoke ~ Sex + Age + Exer,
                            dichotomies = smoke_dichots,
                            data = sv)
summary(smoke_nested)
Anova(smoke_nested)
```

```{r smoke-plot}
plot(smoke_nested, x.var = "Age",
     others = list(Sex = "Female", Exer = "Some"),
     xlab   = "Age",
     main   = "Smoking: Female students, some exercise")
```

**Note**: With only 237 observations and modest predictors, results are
illustrative rather than definitive.

---

# Ordinal examples — brief sketches

## Pneumoconiosis severity (`VGAM::pneumo`)

```{r pneumo-pkg, include=FALSE}
have_VGAM <- requireNamespace("VGAM", quietly = TRUE)
```

```{r pneumo-show, eval=have_VGAM}
data("pneumo", package = "VGAM")
pneumo
```

8 grouped observations on coal miners. Response: `normal` / `mild` / `severe`.
Predictor: `duration` (years of dust exposure).  Very small data but the story
is clean — continuation logits on an ordinal severity scale vs. a single
continuous dose predictor.

```{r pneumo-fit, eval=have_VGAM}
# pneumo is aggregated; nestedLogit() needs individual-level data or
# frequency weights — requires expansion or a weighted interface.
# Placeholder: would use continuationLogits(c("normal", "mild", "severe"))
```

## Mental health status (Agresti)

Mental health status (`Well`, `Mild`, `Moderate`, `Impaired`) vs. SES and
life events — Table 9.1 in Agresti *Categorical Data Analysis* (2002).
May be available via `vcdExtra::Mental` or must be entered from the book.
Classic illustration of continuation logits for an ordinal psychiatric outcome.

---

# Session info

```{r session-info}
sessionInfo()
```
