---
title: "Displaying nestedLogit Models as LaTeX Equations"
author: "Michael Friendly"
date: "`r Sys.Date()`"
package: nestedLogit
output:
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["references.bib", "packages.bib"]
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Displaying nestedLogit Models as LaTeX Equations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message  = FALSE,
  warning  = FALSE,
  comment  = "#>"
)

.opts <- options(digits = 4)

# packages to be cited in packages.bib
.to.cite <- c("nestedLogit", "equatiomatic", "carData")
```

## Overview

The `equatiomatic` package [@R-equatiomatic] provides a general mechanism for
converting fitted statistical models into LaTeX equations, via the function
`extract_eq()`.  For any model class that has a `broom::tidy()` method,
`extract_eq()` can generate both the symbolic form of the model equation and
a version with fitted coefficient values substituted in.

The `nestedLogit` package [@R-nestedLogit] supports `extract_eq()` for objects
of class `"nestedLogit"` through a dedicated S3 method, `extract_eq.nestedLogit`.
Because a nested logit model is represented internally as a collection of binary
logit sub-models — one for each dichotomy — `extract_eq()` generates a separate
equation for each dichotomy and returns them as a named list.

To use these features, load both packages:

```{r setup}
library(nestedLogit)
library(equatiomatic)
```

## Fitting a nested logit model

We use the `Womenlf` data [@R-carData], which records the labour-force
participation of married women and is included in the `carData` package.
The three-category response `partic` (not working, part-time, full-time) is
decomposed into two nested binary dichotomies:

- **work**: not working vs. working (part-time or full-time)
- **full**: part-time vs. full-time, *among those who work*

```{r model}
data(Womenlf, package = "carData")

comparisons <- logits(
  work = dichotomy("not.work", working = c("parttime", "fulltime")),
  full = dichotomy("parttime", "fulltime")
)

wlf.nested <- nestedLogit(partic ~ hincome + children,
                           dichotomies = comparisons,
                           data = Womenlf)
```

## Symbolic equations

Calling `extract_eq()` on a `"nestedLogit"` object returns the model equations
in symbolic (Greek-letter) form, with the response for each dichotomy labelled
by its name.

```{r eqns}
eqns <- extract_eq(wlf.nested)
```

### Work dichotomy (not working vs. working)

```{r eqn-work}
eqns$work
```

### Full-time dichotomy (part-time vs. full-time)

```{r eqn-full}
eqns$full
```

## Equations with fitted coefficients

Passing `use_coefs = TRUE` substitutes the fitted coefficient values into the
equations.

```{r eqns-coef}
eqns_coef <- extract_eq(wlf.nested, use_coefs = TRUE)
```

```{r eqn-coef-work}
eqns_coef$work
```

```{r eqn-coef-full}
eqns_coef$full
```

## Equations for individual sub-models

The individual binary logit sub-models (objects of class `"glm"`) can also be
passed directly to `extract_eq()`.  This can be useful when you want to work
with a single dichotomy in isolation.

```{r submodel}
mod.work <- models(wlf.nested, "work")
extract_eq(mod.work)
```

## Using the raw LaTeX

Each element of the returned list is an object of class `"equation"` (from
`equatiomatic`), which is a character string containing the LaTeX source.
It renders automatically in R Markdown and Quarto documents.  To access
the raw LaTeX — for example to paste it into a paper or to render it with
another tool such as `katex` — use `as.character()`:

```{r raw-latex}
cat(as.character(eqns$work), "\n\n")
cat(as.character(eqns$full), "\n")
```

```{r write-bib, echo=FALSE}
pkgs <- unique(c(.to.cite, .packages()))
knitr::write_bib(pkgs, file = here::here("vignettes", "packages.bib"))
```

## References

```{r, include=FALSE}
options(.opts)
```
