---
title: "Plotting nestedLogit models with ggplot2"
author: "Michael Friendly and John Fox"
date: "`r Sys.Date()`"
package: nestedLogit
output:
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["references.bib", "packages.bib"]
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Plotting nestedLogit models with ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.height = 6,
  fig.width = 7,
  fig.path = "fig/",
  dev = "png",
  comment = "#>"
)

# save some typing
knitr::set_alias(w = "fig.width",
                 h = "fig.height",
                 cap = "fig.cap")

# colorize text
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}


set.seed(47)
.opts <- options(digits = 4)

# packages to be cited here. Code at the end automatically updates packages.bib
#to.cite <- c("ggplot2", "geomtextpath", "equatiomatic")
```

Load the packages we'll use here:
```{r setup}
library(nestedLogit)    # Nested Dichotomy Logistic Regression Models
library(knitr)          # A General-Purpose Package for Dynamic Report Generation in R
library(dplyr)          # A Grammar of Data Manipulation
library(tidyr)          # Tidy Messy Data
library(ggplot2)        # Create Elegant Data Visualisations Using the Grammar of Graphics
library(geomtextpath)   # Curved Text in 'ggplot2'
```

The main vignette illustrated the basic plot method, `plot.nestedLogit()` in the package.
However, for better control of the details and possibly more pleasing graphs,
it is useful to describe how graphs can be constructed using `ggplot2` [@R-ggplot2]. We'll use the
same example of women's labor force participation, using the original dichotomies:

```{r wlf-model}
data(Womenlf, package = "carData")
comparisons <- logits(work=dichotomy("not.work", c("parttime", "fulltime")),
                      full=dichotomy("parttime", "fulltime"))

wlf.nested <- nestedLogit(partic ~ hincome + children,
                          dichotomies = comparisons,
                          data=Womenlf)
```

The advantages of this approach are that 

* you can handle more complicated designs with more predictors by faceting,
* it allows you to plot either the fitted probabilities or their transformed logits,
* you can also obtain plots for log odds corresponding to each of the dichotomies which comprise the nested logit model.

As we will illustrate, this provides a nice visual interpretation of the alternative
specification of dichotomies for the `Womenlf` data discussed in the section
"Alternative models for the `Womenlf` data" of the main vignette.


## Fitted probabilities
To draw a plot, it is sufficient to calculate predicted probabilities over a grid of
values of the predictor variables. Here, we select a range of 0 - 45 in steps of 5,
combined with the two values of `children`.

```{r pred.nested}
new <- expand.grid(hincome=seq(0, 45, by = 5),
                   children=c("absent", "present"))

pred.nested <- predict(wlf.nested, newdata = new)
names(pred.nested)
```

As explained in `help(predict.nestedLogit)`, the predict method returns a complicated structure --
a list of four data frames corresponding to the predicted probabilities for the response categories,
the corresponding logits, and each of their standard errors. 

```{r}
head(pred.nested[["p"]])
```

However, `ggplot` wants the data in long format. 
This is easily done using the `as.data.frame()` method, which also includes the values of the predictors in the `newdata` data set:

```{r}
plotdata <- as.data.frame(pred.nested, newdata=new)
head(plotdata)
```

## Plotting with `ggplot2`
Then we can plot probability against one predictor, use `color` to distinguish the levels of the response
(`partic`) and facet facet the plot by `children`. The point-wise standard errors are drawn
in a 68% confidence envelope using `geom_ribbon()`. We've also plotted the predicted values as points to
show where the predictions are obtained.
```{r wlf-ggplot-p1}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "**ggplot**: Predicted probabilities of working at all or working part time or full time versus husband's income, by `children`"

theme_set(theme_bw(base_size = 14))

gg1 <- ggplot(plotdata,
       aes(x=hincome, y=p, color=response)) +
  geom_line(linewidth = 2) +
  geom_point(size = 1.5, shape = 16, color = "black") +
  labs(x="Husband's Income", y= "Probability") +
  facet_wrap(~ children, labeller = label_both) +
  geom_ribbon(aes(ymin=p - se.p,
                  ymax=p + se.p,
                  fill = response), alpha = 0.3) 

gg1
```
It is noteworthy that the confidence envelopes are wider for not-working women at higher levels of husband's income, where there are fewer observations.

### Direct labels

Plot legends are somewhat hard to read and take up unnecessary space in the plot, so it is often better
to label the curves directly.  The `geomtextpath` package [@R-geomtextpath] produces a nicer plot.

```{r wlf-ggplot-p2}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "The same plot using direct labels on the curves rather than a legend.`"
gg1 + geom_textline(aes(label = response), 
                    hjust = -0.01, vjust=-0.5, size=5) +
  theme(legend.position = "none")
```

### Plotting log-odds
The advantage of the data structure returned by `as.data.frame()` is that you can just as easily plot the predicted probabilities on the scale of log-odds
($\text{logit}(p) = \log(p / 1-p)$),
using the `logit` and
`logit.se` components.

```{r wlf-ggplot-logit}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "Predicted log odds of working at all or working part time or full time versus husband's income, by `children`"

ggplot(plotdata,
       aes(x=hincome, y=logit, color=response)) +
  geom_line(linewidth = 2) +
  geom_point(size = 1.5, shape = 16, color = "black") +
  labs(x="Husband's Income", y= "Log Odds") +
  facet_wrap(~ children, labeller = label_both) +
  geom_ribbon(aes(ymin=logit - se.logit,
                  ymax=logit + se.logit,
                  fill = response), alpha = 0.3) +
  geom_textline(aes(label = response), 
                hjust = -0.01, vjust=-0.5, size=5) +
  theme(legend.position = "none")
```

## Predicted logit values for the dichotomies

The nested logit model `wlf.nested` is comprised of the two binary logistic regression
models for the `work` and `full` dichotomies. We can plot these as follows.

```{r}
names(models(wlf.nested))
```

The `predict()` method can also generate predicted values and their standard errors for the logits of these dichotomies, using the  `model = "dichotomies"` argument:
```{r}
pred.dichot <- predict(wlf.nested, newdata = new,
                       model = "dichotomies")
str(pred.dichot)
```

Transforming this to a data frame, we get an analogous result for plotting:

```{r}
plotlogit <- as.data.frame(pred.dichot, newdata = new)
head(plotlogit)
```

Then, plot the `logit` vs. husband's income, with separate curves for the two
sub-models:

```{r wlf-ggplot-dichot1}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "Predicted logits for the two dichotomies, `work` and `full` versus `hincome`, by `children`"
ggplot(plotlogit,
       aes(x=hincome, y=logit, color=response)) +
  geom_line(linewidth = 2) +
  geom_point(size = 1.5, shape = 16, color = "black") +
  labs(x="Husband's Income", y= "Log Odds") +
  facet_wrap(~ children, labeller = label_both) +
  geom_ribbon(aes(ymin=logit - se.logit,
                  ymax=logit + se.logit,
                  fill = response), alpha = 0.3) +
  geom_textline(aes(label = response),
                hjust = -0.01, vjust=-0.5, size=5) +
  theme(legend.position = "none")
```

Or, interchanging the roles of `children` and `response`, we can plot these the other way, faceting by `response`.

```{r wlf-ggplot-dichot2}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "Predicted logits for the two dichotomies, `work` and `full` versus `hincome`, by `response`"
ggplot(plotlogit,
       aes(x=hincome, y=logit, color=children)) +
  geom_line(linewidth = 2) +
  geom_point(size = 1.5, shape = 16, color = "black") +
  labs(x="Husband's Income", y= "Log Odds") +
  facet_wrap(~ response, labeller = label_both) +
  geom_ribbon(aes(ymin=logit - se.logit,
                  ymax=logit + se.logit,
                  fill = children), alpha = 0.3) +
  geom_textline(aes(label = children),
                hjust = -0.01, vjust=-0.5, size=5) +
  theme(legend.position = "none")
```

This nicely illustrates the nature of the fitted logit models: The lines in each panel
have the same slopes for the two levels of `children`, differing only in their intercepts.
The `full` distinction between working full-time vs. part-time decreases faster with husband's
income than for the `work` dichotomy between not working at all and working either
part-time or full-time.

## Alternative model

In the main vignette we mentioned that
an alternative set of nested dichotomies first contrasts full-time work with the other categories, {full-time} vs. {not working, part-time}, and then {not working} vs. {part-time}.

```{r alt-model}
wlf.nested.alt <- nestedLogit(partic ~ hincome + children,
                              logits(full=dichotomy(nonfulltime=c("not.work", "parttime"), "fulltime"),
                                     part=dichotomy("not.work", "parttime")),
                              data=Womenlf)
```

Proceeding in the same way above, we get predicted logits and standard errors for each
of the dichotomies.

```{r}
pred.dichot.alt <- predict(wlf.nested.alt, newdata = new,
                       model = "dichotomies")
plotlogit.alt <- as.data.frame(pred.dichot.alt, newdata = new)
head(plotlogit.alt)
```

Plotting these as before:
```{r wlf-ggplot-alt1}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "Predicted logits for the two dichotomies, `work` and `full` versus `hincome`, by `response`"
ggplot(plotlogit.alt,
       aes(x=hincome, y=logit, color=children)) +
  geom_line(linewidth = 2) +
  geom_point(size = 1.5, shape = 16, color = "black") +
  labs(x="Husband's Income", y= "Log Odds") +
  facet_wrap(~ response, labeller = label_both) +
  geom_ribbon(aes(ymin=logit - se.logit,
                  ymax=logit + se.logit,
                  fill = children), alpha = 0.3) +
  geom_textline(aes(label = children),
                hjust = -0.01, vjust=-0.5, size=5) +
  theme(legend.position = "none")
```

It's apparent that the alternative model produces a simpler description of the data: The predictors husband's income and presence of children affect the decision to work full-time, but not the decision to work part-time among those who aren't engaged in full-time work.
In particular it is clear that neither husband's income nor having young children has any effect on the
decision to work part-time.

```{r, include = FALSE}
options(.opts)
```

## References


