---
title: "Plotting nestedLogit models"
author: "Michael Friendly and John Fox"
date: "`r Sys.Date()`"
package: nestedLogit
output:
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["references.bib", "packages.bib"]
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Plotting nestedLogit models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.height = 6,
  fig.width = 7,
  fig.path = "fig/",
  dev = "png",
  comment = "#>"
)

# save some typing
knitr::set_alias(w = "fig.width",
                 h = "fig.height",
                 cap = "fig.cap")

# colorize text
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}


set.seed(47)
.opts <- options(digits = 4)

# packages to be cited here. Code at the end automatically updates packages.bib
to.cite <- c("ggplot2", "geomtextpath", "equatiomatic")
```

Load the packages we'll use here:
```{r setup}
library(nestedLogit)    # Nested Dichotomy Logistic Regression Models
library(knitr)          # A General-Purpose Package for Dynamic Report Generation in R
library(dplyr)          # A Grammar of Data Manipulation
library(tidyr)          # Tidy Messy Data
library(ggplot2)        # Create Elegant Data Visualisations Using the Grammar of Graphics
library(geomtextpath)   # Curved Text in 'ggplot2'
```

The main vignette illustrated the basic plot method, `plot.nestedLogit()` in the package.
However, to explain plotting nested-dichotomies models works, and for better control of the details,
it is useful to describe how graphs can be constructed directly directly. We'll use the
same example of women's labor force participation, using the original dichotomies:

```{r wld-model}
data(Womenlf, package = "carData")
Womenlf$partic <- with(Womenlf,
                       factor(partic, levels = c("not.work", "parttime", "fulltime")))

wlf.nested <- nestedLogit(partic ~ hincome + children,
                          dichotomies = logits(work=dichotomy("not.work", working=c("parttime", "fulltime")),
                                               full=dichotomy("parttime", "fulltime")),
                          data=Womenlf)
```

## Fitted probabilities
To draw a plot, it is sufficient to calculate predicted probabilities over a grid of
values of the predictor variables. Here, we select a range of 0 - 45 in steps of 5,
combined with the two values of `children`.

```{r plotdata}
new <- expand.grid(hincome=seq(0, 45, by = 5),
                   children=c("absent", "present"))

pred.nested <- predict(wlf.nested, newdata = new)
plotdata <- cbind(new, pred.nested)
head(plotdata)
```

### Using `matplot()`
Because the fitted values are in multiple columns, it is easiest to plot these using
`matplot()`. We could plot these all in a single (messy) figure, but it is clearer
to show separate panels for children `absent` and `present`.

To do this, the `plotdata` data set is subset within a loop over the values of `children`
and each subset is plotted by `matplot`. It is only necessary to include the legend
in one panel. The plots are combined into a single figure using `par(mfrow())`.
```{r wlf-matplot}
#| fig.show = "hold",
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "**matplot**: Predicted probabilities of working at all or working part time or full time"
op <- par(mfrow=c(1,2), mar=c(4,4,3,1)+.1)
cols=c("blue", "magenta", "darkgreen")
for ( kids in c("absent", "present") ) {
  data <- subset(plotdata, children==kids)
  matplot(data[, "hincome"], data[, 5:3], 
          type = "l", lwd=3, lty = 1:3, col = cols,
          xlab="Husband's Income", 
          ylab='Fitted Probability',
          main = paste("Children", kids),
          cex.lab = 1.1)
  if (kids=="absent") {
    legend("topright", lty=1:3, lwd=3, col=cols, bty = "n",
           legend=c("fulltime", "parttime", "not working"))
  }
}
par(op)
```

### Using `ggplot()`

More control, and perhaps a more aesthetically pleasing figure can be produced using `ggplot`
[@R-ggplot2].
However, `ggplot` wants the data in long format. That makes it easy to plot probability
against one predictor and use `color` to distinguish the levels of
`partic` and facet the plot by `children`.
(The result of `tidyr::pivot_longer()` doesn't recognize an ordered nature of `"Working"`,
so this is done in a separate step.)


```{r plotlong}
plotlong <- plotdata |>
  tidyr::pivot_longer(fulltime : not.work,
                      names_to = "Working",
                      values_to = "Probability") |>
  mutate(Working = ordered(Working, 
                           levels = c("not.work", "parttime", "fulltime")) )

head(plotlong)
```

Then, one call to `ggplot` produces both panels. To sort the levels of `Working` appropriately,
we made this an ordered factor in the step above, but assign the colors with a discrete scale.
```{r wlf-ggplot}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "**ggplot**: Predicted probabilities of working at all or working part time or full time"

gg <- ggplot(plotlong,
             aes(x=hincome, y=Probability, color=Working)) +
  geom_line(linewidth = 2) +
  scale_color_discrete() +
  labs(x="Husband's Income", y= "Probability") +
  facet_wrap(~ children, labeller = label_both) +
  theme_bw(base_size = 14) +
  theme(legend.position = c(.3, .8))
gg
```

### Direct labels

It's usually nicer to label the curves directly than to rely on a legend.
The `geomtextpath` [@R-geomtextpath] package offers `geom_textline()` as an alternative
to `geom_line()` that adds a text label to a curve. `hjust` and `vjust` position the labels
along the curve.

```{r wlf-geomtextpath}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "**geomtextpath**: Predicted probabilities, with labels on the curves"
ggplot(plotlong,
       aes(x=hincome, y=Probability, color=Working)) +
  geom_textline(aes(label = Working),
                linewidth = 2, size = 5, 
                hjust = 0.9, vjust = 0.2) +
  scale_color_discrete() +
  labs(x = "Husband's Income", y = "Probability") +
  facet_wrap(~ children, labeller = label_both) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")
```

## Plotting fitted log odds

<!-- Michael: Maybe move this to a separate vignette? -->


It is sometimes easier to interpret logistic regression models by plotting the linear predictors
on the **log odds** scale, because these appear as straight lines. That is, from the estimated
coefficients in the model, the predicted log odds of `work` and `full` are given by
<!-- TODO: cite equatiomatic -->
<!-- equatiomatic::extract_eq(wlf.nested$models[[1]], use_coefs = TRUE) -->
<!-- equatiomatic::extract_eq(wlf.nested$models[[2]], use_coefs = TRUE) -->

$$
L_1 =\log\left[ \frac { \widehat{P( \operatorname{work} = \operatorname{1} )} }{ 1 - \widehat{P( \operatorname{work} = \operatorname{1} )} } \right] = 1.34 - 0.04(\operatorname{hincome}) - 1.58(\operatorname{children}_{\operatorname{present}})
$$

$$
L_2 =\log\left[ \frac { \widehat{P( \operatorname{full} = \operatorname{1})} }{ 1 - \widehat{P( \operatorname{full} = \operatorname{1})} } \right] = 3.48 - 0.11(\operatorname{hincome}) - 2.65(\operatorname{children}_{\operatorname{present}})
$$
Here, $\operatorname{children}_{\operatorname{present}}$ is either 0 or 1, so this
amounts to a shift in the intercept when children are present.

These values can be calculated by calling `predict.glm()` directly for the
`models` component of `wlf.nested`, specifying `type = "link"`:

```{r pred-logits}
pred.logits <- sapply(models(wlf.nested), predict, newdata=new, type = "link")
plotdatal <- cbind(new, pred.logits)
head(plotdatal)
```

Then we plot these more or less as before:
```{r wlf-logits}
#| fig.show = "hold",
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "**log odds**: Predicted logits of the  `work` and `full` dichotomies, by `children`"
cols <- c("blue", "red")

op <- par(mfrow=c(1,2), mar=c(4,4,3,1)+.1)
for ( kids in c("absent", "present") ) {
  data <- subset(plotdatal, children==kids)
  matplot(data[, "hincome"], data[, 3:4],
          type = "l", lwd=3, lty = 1, col = cols,
          xlab="Husband's Income",
          ylab='Predicted Log Odds',
          main = paste("Children", kids),
          cex.lab = 1.1)
  if (kids=="absent") {
    legend("topright", lty=1, lwd=3, col=cols, bty = "n",
           title = "Dichotomy",
           legend=c("work", "full"))
  }
}
par(op)
```

Finally, we can use `ggplot` as before to plot the fitted logits, first transforming the long format.

```{r plotlongl}
plotlongl <- plotdatal |>
  tidyr::pivot_longer(work : full,
                      names_to = "Dichotomy",
                      values_to = "logit") |>
  mutate(Dichotomy = ordered(Dichotomy,
                         levels = c("work", "full")) )
```

Doing this gives the flexibility that we can facet the plot either by `children` (as before)
or `Dichotomy`. The latter nicely shows that the additive model has equal slopes for husband's
income within both panels: increasing husband's income decreases the log odds of working,
but at a faster rate for the `full` dichotomy.
Having young children decreases the log odds of working either at all
or working fulltime as compared to parttime.
```{r wlf-gglogits}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "**log odds**: Predicted logits of the  `work` and `full` dichotomies, by `Dichotomy`"
ggplot(plotlongl,
       aes(x=hincome, y=logit, color=children)) +
  geom_line(linewidth = 3) +
  geom_point(size = 1.5, shape = 16, color = "black") +
  scale_color_discrete() +
  labs(x="Husband's Income", y= "Log Odds") +
  facet_wrap(~ Dichotomy, labeller = label_both) +
  theme_bw(base_size = 14) +
  theme(legend.position = c(.35, .82))

```

<!-- Can't do it this way as that would clobber the one written by the other vignette. -->

```{r write-bib, echo = FALSE}
# write a packages.bib file of the packages (.packages()) that have been used here
#pkgs <- unique(c(to.cite, .packages()))
#knitr::write_bib(pkgs, file = here::here("vignettes", "packages.bib"))
```

```{r, include = FALSE}
options(.opts)
```

## References


