---
title: "Using easystats with nestedLogit models"
author: "Michael Friendly"
date: "`r Sys.Date()`"
package: nestedLogit
output:
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["../references.bib", "../packages.bib"]
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Using easystats with nestedLogit models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.height = 6,
  fig.width = 7,
  fig.path = "fig/",
  dev = "png",
  comment = "#>"
)

# save some typing
knitr::set_alias(w = "fig.width",
                 h = "fig.height",
                 cap = "fig.cap")

set.seed(47)
.opts <- options(digits = 4)
```

## Overview

The [easystats](https://easystats.github.io/easystats/) project [@easystats2022] is a collection
of R packages that provides a unified, consistent framework for working with statistical models.
Three of its core packages offer direct support for `"nestedLogit"` objects:

- **insight** [@R-insight]: Extracts model information --- formulas, data, parameters, variance-covariance
  matrices --- through a unified interface.
- **parameters** [@R-parameters]: Computes and formats model parameter tables with confidence intervals,
  standardized coefficients, and bootstrap estimates.
- **performance** [@R-performance]: Calculates model performance indices such as R-squared measures
  and fit statistics.

These packages work by iterating over the individual binary logit sub-models that
comprise a nested dichotomy model, combining the results into tidy tables with a
`Response` column identifying each dichotomy. This means you get a unified view
of all dichotomies without having to process them individually.

Additionally, the `see` package provides `plot()` methods for the output of
`parameters` and `performance`, enabling visual summaries like forest plots.

This vignette describes some ways to enhance what is available in the [nestedLogit](https://friendly.github.io/nestedLogit/)
package using [easystats](https://easystats.github.io/easystats/) capabilities.

Load the packages we'll use here:
```{r setup}
library(nestedLogit)    # Nested Dichotomy Logistic Regression Models
library(insight)        # Easy Access to Model Information
library(parameters)     # Processing of Model Parameters
library(performance)    # Assessment of Regression Models Performance
library(see)            # Visualisation Toolbox for easystats
library(ggplot2)        # Data Visualisations Using the Grammar of Graphics
```

## Women's labor force participation

We use the standard `Womenlf` example. The response `partic` has three categories ---
not working, working part-time, and working full-time --- modeled as nested dichotomies
against husband's income and presence of young children.

```{r wlf-model}
data(Womenlf, package = "carData")
comparisons <- logits(work = dichotomy("not.work", c("parttime", "fulltime")),
                      full = dichotomy("parttime", "fulltime"))

wlf.nested <- nestedLogit(partic ~ hincome + children,
                          dichotomies = comparisons,
                          data = Womenlf)
```

## Model information with `insight`

The `insight` package provides accessor functions that work uniformly across many
model types. For `nestedLogit` models, these extract information from the overall
model and, where applicable, from each dichotomy sub-model.

### Model type and formula

`model_info()` retrieves information about a given model. The model family (binomial) and link function (logit)
are recorded, as well as various `is_*` properties.

```{r insight-model-info}
info <- insight::model_info(wlf.nested)

names(info)
```

The relevant values here are:
```{r insight-model-info2}
info$is_binomial
info$family
info$link_function
```

`find_formula()` returns the model formula. With `dichotomies = TRUE`, it returns
the formula for each sub-model:

```{r insight-formula}
insight::find_formula(wlf.nested)
insight::find_formula(wlf.nested, dichotomies = TRUE)
```

### Parameters and data

`find_parameters()` lists all parameter names across the dichotomies:

```{r insight-params}
insight::find_parameters(wlf.nested)
```

`get_varcov()` returns the variance-covariance matrix for each sub-model.
The `component` argument lets you select a specific dichotomy:

```{r insight-vcov}
insight::get_varcov(wlf.nested)
```


## Model parameters with `parameters`

### Basic parameter table

The `model_parameters()` function produces a formatted summary of all coefficients
across the dichotomies. The printed output includes estimates, standard errors, confidence
intervals, and test statistics. The data structure contains a `Response` column identifying which dichotomy
each row belongs to:

```{r params-basic}
mp <- model_parameters(wlf.nested)
mp
```

### Plot of coefficients

The `see` package provides a `plot()` method for parameter tables, producing a
forest plot of the coefficient estimates with confidence intervals.
This is intended to give a visual comparison of the effects across the two dichotomies.

There is a small incompatibility to work around: 

* `see::plot.parameters_model()` internally passes `Response` column values to `gsub()` as regular expression patterns.
* The curly-brace notation that `nestedLogit` uses to label multi-category dichotomy
sides (e.g., `{parttime} vs. {fulltime}`) contains `{` and `}` characters that are
invalid regex quantifiers, causing an error. 
* Stripping the braces before plotting resolves this:

```{r params-forest, out.width="100%", fig.height=5, fig.cap="Forest plot of model coefficients from `model_parameters()`."}
mp_plot <- mp
mp_plot$Response <- gsub("[{}]", "", mp_plot$Response)
plot(mp_plot)
```

Note that the two predictors are on very different scales: `hincome` is a continuous
variable (measured in thousands of dollars) while `children` is binary (present/absent).
This makes direct visual comparison of their coefficient sizes within the same panel
potentially misleading. The `model_parameters()` function offers a `standardize = "refit"`
argument that re-fits the model on z-scored predictors, which would put all coefficients
on a common standard-deviation scale. 

Unfortunately, this option does not currently
produce a visibly different result for `nestedLogit` objects, suggesting that
standardization is not yet fully propagated through the nested structure. Until this
is resolved, the `compare_parameters()` approach in the next section --- which works
directly on the individual `glm` sub-models --- is more reliable for visual comparison.

### Comparing dichotomies

`compare_parameters()` places the estimates from different sub-models side by side,
making it easy to see how the same predictors operate differently across the
nested dichotomies. Here we extract the individual `glm` sub-models using
`models()` and compare them:

```{r params-compare}
cp <- compare_parameters(models(wlf.nested, "work"),
                         models(wlf.nested, "full"),
                         column_names = c("Work", "Full"))
cp
```

There is a plot method for the result, which plots the coefficients for the two sup-models
in a single plot.

```{r params-compare-plot, out.width="100%", fig.height=5, fig.cap="Comparison of coefficients across the two dichotomy sub-models."}
plot(cp)
```

### Bootstrap confidence intervals

`model_parameters()` supports bootstrap-based inference. This re-fits the model
on resampled data to produce non-parametric bootstrap confidence intervals,
which can be more robust than the default Wald intervals for smaller samples:

```{r params-bootstrap}
model_parameters(wlf.nested, bootstrap = TRUE, iterations = 500)
```


## Model performance with `performance`

### Performance summary

`model_performance()` provides a one-line summary of fit statistics for each
dichotomy sub-model, including AIC, BIC, Tjur's R-squared, RMSE, and
log-likelihood:

```{r perf-summary}
model_performance(wlf.nested)
```

### R-squared measures

Several R-squared variants are available for logistic regression models.
Tjur's R-squared (also known as the coefficient of discrimination) is the
default for binomial models:

```{r perf-r2}
r2(wlf.nested)
```

Other R-squared measures provide alternative perspectives on model fit:

```{r perf-r2-other}
r2_nagelkerke(wlf.nested)
r2_coxsnell(wlf.nested)
```


## Diagnostics for individual dichotomies

While the easystats functions above work on the `nestedLogit` object as a whole,
some diagnostic tools (such as `check_model()`) require a standard `glm` object.
The `models()` function extracts individual dichotomy sub-models, which are
ordinary `glm` objects that work with the full range of easystats diagnostic tools.

### Extracting sub-models

```{r submodel-extract}
wlf_work <- models(wlf.nested, "work")  # working vs. not working
wlf_full <- models(wlf.nested, "full")  # full-time vs. part-time

class(wlf_work)
```

### Visual diagnostics with `check_model()`

`performance::check_model()` produces a comprehensive set of diagnostic plots
for a `glm` object, including checks for influential observations, collinearity,
and the distribution of residuals. 

<!--
Here, I only show the influence plot 
(poorly named: `check = "outliers"`).

I get error when I only try to plot the outliers/influence plot
check_model(wlf_work, check = "outliers", residual_type = "normal")
-->

```{r diag-work, out.width="100%", fig.height=8, fig.cap="Diagnostic plots for the work vs. not-work dichotomy."}
check_model(wlf_work)
```

```{r diag-full, out.width="100%", fig.height=8, fig.cap="Diagnostic plots for the full-time vs. part-time dichotomy."}
check_model(wlf_full)
```

### Binned residual plot

For binomial models, `performance::binned_residuals()` provides a binned
residual plot, a more appropriate residual diagnostic than raw residuals.
Points falling outside the error bounds suggest potential model misfit.

`binned_residuals()` requires the binary response variable to be present as a
named column in the model's data. The sub-models returned by `models()` have
their `call$data` reset to the original dataset (`Womenlf`), which does not
contain the binary response columns --- those were created internally as a
temporary variable during model fitting and were never added back to the
original data. As a result, `insight::get_data()` cannot find the response
column and errors.

The workaround is to construct small data frames with the binary response
defined explicitly, matching the logic used by `nestedLogit` internally:

```{r diag-binned-work, out.width="100%", fig.height=5, fig.cap="Binned residual plot for the work vs. not-work dichotomy."}
# work dichotomy: all cases, not.work = 0, parttime or fulltime = 1
df_work <- within(Womenlf, work <- as.integer(partic != "not.work"))
glm_work2 <- glm(work ~ hincome + children, data = df_work, family = binomial)
check_work <- binned_residuals(glm_work2)
plot(check_work)
```

```{r diag-binned-full, out.width="100%", fig.height=5, fig.cap="Binned residual plot for the full-time vs. part-time dichotomy."}
# full dichotomy: workers only, parttime = 0, fulltime = 1
df_full <- subset(Womenlf, partic != "not.work")
df_full$full <- as.integer(df_full$partic == "fulltime")
glm_full2 <- glm(full ~ hincome + children, data = df_full, family = binomial)
check_full <- binned_residuals(glm_full2)
plot(check_full)
```


## Limitations

The easystats ecosystem provides substantial support for `nestedLogit` models
through `insight`, `parameters`, and `performance`. However, some packages
in the ecosystem do not currently support `nestedLogit`:

- **modelbased**: Functions like `estimate_expectation()` and `estimate_prediction()`
  do not work with `nestedLogit` objects. For predicted probabilities and marginal
  effects, use the `ggeffects` package instead (see
  `vignette("ggeffects", package = "nestedLogit")`).

- **Diagnostic plots**: `check_model()` does not work directly on `nestedLogit`
  objects, but as shown above, it can be applied to the individual `glm`
  sub-models extracted with `models()`.

For plotting predicted probabilities of the response categories, see
`vignette("ggeffects", package = "nestedLogit")`. For fully customized
`ggplot2` plots including the dichotomy-level predictions, see
`vignette("plotting-ggplot", package = "nestedLogit")`.

## References

```{r, include = FALSE}
options(.opts)
```
