---
title: "Using ggeffects with nestedLogit models"
author: "Michael Friendly and John Fox"
date: "`r Sys.Date()`"
package: nestedLogit
output:
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["../references.bib", "../packages.bib"]
csl: ../apa.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.height = 6,
  fig.width = 7,
  fig.path = "fig/",
  dev = "png",
  comment = "#>"
)

# save some typing
knitr::set_alias(w = "fig.width",
                 h = "fig.height",
                 cap = "fig.cap")

set.seed(47)
.opts <- options(digits = 4)
```

Load the packages we'll use here:
```{r setup}
library(nestedLogit)    # Nested Dichotomy Logistic Regression Models
library(ggeffects)      # Create Tidy Data Frames of Marginal Effects
library(ggplot2)        # Create Elegant Data Visualisations Using the Grammar of Graphics
```

## Overview

The `ggeffects` package [@R-ggeffects; @ggeffects2018] provides a simple and unified interface for
computing and plotting adjusted predictions and marginal effects from a wide
variety of regression models. Its main function, `predict_response()`,
returns a tidy data frame of model predictions that can be plotted directly
with a built-in `plot()` method or further customized with `ggplot2`.

The package now supports `"nestedLogit"` objects, making it easy to visualize predicted
probabilities for each response category across levels of the predictors, without the
manual data wrangling described in `vignette("plotting-ggplot", package = "nestedLogit")`.

## Women's labor force participation

We use the standard `Womenlf` example from the main vignette. The response `partic` has
three categories --- not working, working part-time, and working full-time --- modeled
as nested dichotomies against husband's income and presence of young children.

```{r wlf-model}
data(Womenlf, package = "carData")
comparisons <- logits(work = dichotomy("not.work", c("parttime", "fulltime")),
                      full = dichotomy("parttime", "fulltime"))

wlf.nested <- nestedLogit(partic ~ hincome + children,
                          dichotomies = comparisons,
                          data = Womenlf)
```

## Predicted probabilities with `predict_response()`

The simplest way to obtain predicted probabilities and a plot is with
`predict_response()`, specifying the focal predictors in the `terms` argument.
This returns predicted probabilities for each response level, with confidence intervals,
averaged over the non-focal predictors.

```{r wlf-predict-response}
wlf.pred <- predict_response(wlf.nested, terms = c("hincome", "children"))
wlf.pred
```

The default `plot()` method produces a panel for each response category:

```{r wlf-ggeffects-plot1}
#| out.width = "100%",
#| fig.height = 5,
#| fig.cap = "Predicted probabilities from `predict_response()` with default plot."
plot(wlf.pred)
```

## Customizing the plot

The `plot()` method returns a `ggplot` object, so it can be further customized with
standard `ggplot2` functions. For example, we can adjust the line size, labels,
theme, and legend position:

```{r wlf-ggeffects-plot2}
#| out.width = "100%",
#| fig.height = 5,
#| fig.cap = "Customized `ggeffects` plot with adjusted labels and theme."
plot(wlf.pred,
     line_size = 2) +
  labs(title = "Predicted Probabilities of Work by Husband's Income",
       y = "Probability",
       x = "Husband's Income") +
  theme_ggeffects(base_size = 16) +
  theme(legend.position = "top")
```

## Plotting on the logit scale

`ggplot2` provides a built-in `"logit"` transformation for axes via
`scale_y_continuous(transform = "logit")`. This displays predicted probabilities
on the logit scale, $\text{logit}(p) = \log(p / (1-p))$, where the axis labels
remain as probabilities but their spacing reflects the logit transformation.
This is useful because the logistic regression model is linear on the logit scale,
so the predicted curves appear as straighter lines.

Since the `plot()` method for `ggeffects` returns a `ggplot` object, we can
simply add this scale transformation. Note that we need to specify breaks
manually, because the automatic break algorithm does not work well with the
logit transformation.

```{r wlf-logit-scale}
#| out.width = "100%",
#| fig.height = 5,
#| fig.cap = "Predicted probabilities on the logit scale."
plot(wlf.pred,
     line_size = 2) +
  scale_y_continuous(
    transform = "logit",
    breaks = c(0.05, 0.10, 0.25, 0.50, 0.75, 0.90, 0.95)
  ) +
  labs(title = "Predicted Probabilities (logit scale)",
       y = "Probability (logit scale)",
       x = "Husband's Income") +
  theme_ggeffects(base_size = 16) +
  theme(legend.position = "top")
```

## Alligator food choice

As a simpler example with a single continuous predictor, we fit a nested logit model to the
`gators` data, predicting primary food choice from alligator length.
The first dichotomy contrasts {Other} vs. {Fish, Invertebrates}, and the second
contrasts {Fish} vs. {Invertebrates}.

```{r gators-model}
data(gators)
gators.nested <- nestedLogit(food ~ length,
                             dichotomies = logits(
                               other   = dichotomy("Other", c("Fish", "Invertebrates")),
                               fish_inv = dichotomy("Fish", "Invertebrates")),
                             data = gators)
```

```{r gators-ggeffects}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "Predicted food choice probabilities for alligators by length."
predict_response(gators.nested, terms = "length") |> 
  plot(line_size = 2)
```

As you can see the main thing going on here is that larger alligators prefer fish, which smaller ones prefer invertebrates.

For comparison, the basic `nestedLogit::plot()` method using default arguments gives a similar plot, with the three curves 
overlaid in a single panel (it uses `graphics::matplot()`).

```{r gators-plot}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "Predicted food choice probabilities for alligators by length."
plot(gators.nested, x.var = "length", 
     legend.bty = "o")
```

## Limitations

The `ggeffects` package computes and plots predicted probabilities for the
_response categories_ of a nested logit model. As shown above, these can be
displayed on the logit scale using `ggplot2`'s built-in axis transformation.

However, `ggeffects` does not currently provide access to the individual
dichotomy sub-models that comprise the nested logit --- for example, plotting
predicted values for the `work` and `full` dichotomies separately.

For these more specialized displays, see `vignette("plotting-ggplot", package = "nestedLogit")`,
which describes a manual workflow using `predict()` with `model = "dichotomies"` and
`as.data.frame()` to construct fully customized `ggplot2` plots.

## References

```{r, include = FALSE}
options(.opts)
```
